---
title: "Reproduction of Tutorial: Statistical analysis and reporting of clinical pharmacokinetic studies"
author: "GT"
format: html
editor: visual
editor_options: 
  chunk_output_type: console
---

## Libraries

```{r}
library(readxl)
library(reshape2)
library(tidyverse)  
library(ncappc)
library(Publish) 
library(ggpubr) 
library(ggthemes)  
library(RColorBrewer)  
library(tmaptools)  
library(skimr)
library(styler)
library(pwr)

```


## Load Raw Data

```{r}
rawpk <- readxl::read_xlsx("Tutorial- Statistical analysis and reporting of clinical pharmacokinetic studies Table S1.xlsx")
head(rawpk)
skim(rawpk)
```


## Section I: Plotting a plasma concentration-time curve for Figure 2
```{r}
# Generate a new dataset with columns for mean concentration and standard deviation (SD).
ctdata <- rawpk %>% 
  group_by(TAD, IND) %>%
  dplyr::summarise(
    mean_conc = mean(DV, na.rm = TRUE), 
    SD_conc = sd(DV, na.rm = TRUE)
    )
ctdata %>% print(n = 10^4)

# Recode IND from numeric to character.
ctdata <- ctdata %>% 
  dplyr::mutate(IND = dplyr::recode(IND, "0" = "No inducer", "1" = "Inducer"))
ctdata %>% print(n = 10^4)

# Plotting a simple graph of mean midazolam plasma concentration and time.
ggplot() + 
  geom_point(data = ctdata, aes(x=TAD, y=mean_conc, group=IND))

# Plotting mean midazolam plasma concentration (and standard deviation (SD)) and time.
ctdata %>% 
  ggplot() +
  geom_point(aes(x=TAD, y=mean_conc, group=IND, color=IND)) +
  geom_line(aes(x=TAD, y=mean_conc, group=IND, color=IND)) +
  geom_errorbar(
    aes(
      x=TAD, y = mean_conc, 
      ymax = mean_conc + SD_conc, ymin = mean_conc - SD_conc, 
      color=IND
    ),
    position = position_dodge(width=0.22)
    ) +
    labs(
      x ="Time after dose (h)", 
      y = "Concentration (ng/ml)", 
      title ="Mean midazolam plasma concentrations", 
      color = ""
    ) +
    theme_classic()

# Create y-log scale graph for inset in the main graph.
log_plot <- ggplot(data = ctdata, aes(x = TAD, y = mean_conc, group = IND)) +
  geom_point(aes(color = IND, shape = IND)) +
  geom_line(aes(linetype = IND, color = IND)) +
  labs(x = "Time after dose (h)", y = "Concentration\n (ng/ml)") +
  scale_color_brewer(palette = "Set2") +
  theme_clean() +
  theme(plot.title = element_text(hjust = 2), legend.position = "none", 
        text=element_text(size=12)) +
  scale_x_continuous(limits=c(0, 12), breaks=c(0, 4, 8, 12)) +
  scale_y_log10() +
  theme(text=element_text(family="Times"))

# Combine the main graph and the y-log scale graph in one graph. 
ggplot(data = ctdata, aes(x = TAD, y = mean_conc, group = IND)) +
  geom_point(aes(color = IND, shape = IND)) +
  geom_line(aes(linetype = IND, color = IND)) +
  geom_errorbar(aes(ymax = mean_conc + SD_conc, ymin = mean_conc - SD_conc, color = IND), 
  position=position_dodge(width = .22)) +
  labs(
    x = "Time after dose (h)", 
    y = "Concentration (ng/ml)", 
    title = "Mean midazolam plasma concentration", 
    shape = "Induction", color = "Induction", linetype = "Induction"
  ) +
  scale_color_brewer(palette = "Set2") +
  theme_clean() +
  theme(
    plot.title = element_text(hjust = 0.5), 
    legend.position = "bottom", 
    legend.title = element_blank(), 
    legend.text=element_text(family="Times"), 
    axis.text=element_text(size=12), 
    axis.title=element_text(size=14),
    text=element_text(family="Times")
  ) +
  annotation_custom(
    ggplotGrob(log_plot), 
    xmin = 6.5, xmax = 12.8, 
    ymin = 0.6, ymax = 1.45
  )
```


## Section II: Non-compartmental pharmacokinetic analysis

```{r}
# Compute the non-compartmental pharmacokinetic analysis.
ncappc(
  obsFile = rawpk,
  str1Nm = "IND",
  str1 = c(0,1),
  concUnit = "ng/ml",
  timeUnit = "hours",
  doseUnit = "ng",
  idNmObs = "ID",
  timeNmObs = "TAD",
  concNmObs = "DV",
  doseAmtNm = "DOSE",
  onlyNCA = TRUE,
  LambdaTimeRange = "NULL",
  method = "linearup-logdown",
  studyName = "PK Tutorial",
  outFileNm = "WOW_PK_TUTORIAL",
  extrapolate = TRUE,
  gg_theme = theme_classic()
)

# Create dataframe containing only (Cmax, AUClast, HL_lambda_z (terminal half-life), Cl_obs).
pk_analysis <- ncaOutput %>%
  select(ID, IND, Tmax, Cmax, Cl_obs, HL_Lambda_z, AUClast)

# Change the unit of Cl_obs from ml/h to l/h, to increase readability of the coming Table 2. 
pk_analysis <- pk_analysis %>% 
  mutate(Cl_obs = Cl_obs/1000)
```

## Section III: Descriptive stats for Table 2 (NCP analysis stratified by DDI)

```{r}
# Reshape data from long format to wide format.
pk_analysis_wide <- pk_analysis  %>%
  pivot_longer(
    cols = !c(ID, IND),  # Columns to pivot (all except ID and IND)
  )  %>% 
  pivot_wider(
    names_from = c(name, IND),  # Combine `variable` and `IND` to form new column names
    values_from = value            # Fill these new columns with the `value` column
  )

# Add a new column with ratio.
pk_analysis_wide <- pk_analysis_wide %>%
  mutate(
    Cmax_ratio = Cmax_1 / Cmax_0,
    Cl_obs_ratio = Cl_obs_1 / Cl_obs_0,
    HL_Lambda_z_ratio = HL_Lambda_z_1 / HL_Lambda_z_0,
    AUClast_ratio = AUClast_1 / AUClast_0
  )

# Define the function for calculating geometric mean ratio and 95% CI.
geo <- function(x) {
  aggr <- ci.mean(x, statistic = "geometric")
  paste0(
    round(aggr$geomean, 2), " (", 
    round(aggr$lower, 2), "-", 
    round(aggr$upper, 2), ")"
  )
}

# Define the function for calculating median.
medians <- function(x) {
  med <- median(x)
  iqr_low <- quantile(x, 0.25)
  iqr_up <- quantile(x, 0.75)
  paste0(round(med, 2), " (", round(iqr_low, 2), "-", round(iqr_up, 2), ")")
}

# Calculate geometric mean and 95% CI for the individual parameters.
Cmax_GMR <- geo(pk_analysis_wide$Cmax_ratio)
Cl_GMR <- geo(pk_analysis_wide$Cl_obs_ratio)
HL_GMR <- geo(pk_analysis_wide$HL_Lambda_z_ratio)
AUC_GMR <- geo(pk_analysis_wide$AUClast_ratio)
Tmax_GMR <- "NA"

# Assemble results in one column for Table 2.
results_GMR <- rbind(AUC_GMR, Cmax_GMR, HL_GMR, Cl_GMR, Tmax_GMR)

# Calculate median for the individual parameters with and without inducer.
Cmax_0_M <- medians(pk_analysis_wide$Cmax_0)
Cmax_1_M <- medians(pk_analysis_wide$Cmax_1)
Cl_0_M <- medians(pk_analysis_wide$Cl_obs_0)
Cl_1_M <- medians(pk_analysis_wide$Cl_obs_1)
HL_0_M <- medians(pk_analysis_wide$HL_Lambda_z_0)
HL_1_M <- medians(pk_analysis_wide$HL_Lambda_z_1)
AUC_0_M <- medians(pk_analysis_wide$AUClast_0)
AUC_1_M <- medians(pk_analysis_wide$AUClast_1)
Tmax_0_M <- medians(pk_analysis_wide$Tmax_0)
Tmax_1_M <- medians(pk_analysis_wide$Tmax_1)

# Assemble results in two columns (with and without induction) for Table 2.
results_IND <- rbind(AUC_1_M, Cmax_1_M, HL_1_M, Cl_1_M, Tmax_1_M)
results_noIND <- rbind(AUC_0_M, Cmax_0_M, HL_0_M, Cl_0_M, Tmax_0_M)

# Assemble Table 2
Table2 <- matrix(nrow=5, ncol=0)
Table2 <- cbind(results_noIND, results_IND, results_GMR, Table2)
colnames1 <- c("WITHOUT INDUCER (Median (IQR))", "WITH INDUCER (Median (IQR))", "GMR (95% CI)")
Table2 <- rbind(colnames1, Table2)

rownames1 <- c("DRUG", "Midazolam", "", "", "", "")
rownames2 <- c("PARAMETER", "AUC 0-last (ng*h ml^-1)", "Cmax (ng ml^-1)", "T1/2 (h)", "CL/F (L h^-1)", "Tmax (h)")
Table2 <- cbind(rownames1, rownames2, Table2)

# Test Tmax with paired Wilcoxon rank-sum test. 
Test_tmax <- wilcox.test(ncaOutput$Tmax, ncaOutput$IND, paired = TRUE)

```

## Section IV: Plotting individual spaghettiplots for Figure 2

```{r}
# Recode the induction variable from numeric to categorical.
Data_fig2 <- ncaOutput
Data_fig2$IND <- factor(ncaOutput$IND, levels= c("0", "1"), labels=c("No inducer", "Inducer"))

# Change the unit of Cl_obs from ml/h to l/h, to increase readability of the coming Table 2. 
Data_fig2$Cl_obs <- (Data_fig2$Cl_obs)/1000

# Create a spaghettiplot for the individual clearances. 
ggplot(data = Data_fig2, aes(x = IND, y = Cl_obs)) +
  geom_point(size=1) +
  geom_line(aes(group = ID), size=0.5) + 
  theme_classic() +
  labs(x = "", y = expression(Midazolam~CL/F~(l~h^-1))) +
  scale_y_continuous(position = "left", breaks = scales ::pretty_breaks(n=4)) +
  guides(color=none) +
  theme(text=element_text(family="Times"))
```


## Section V: Concentration-time curve for individual data points

```{r}
# Recode ID from numeric to categorical.
rawpk$ID <- factor(rawpk$ID)

# Recode IND from numeric to character.
rawpk$IND <- recode(rawpk$IND, "0" = "No inducer", "1" = "Inducer")

# Plot midazolam plasma concentrations and time for the individual subjects. 
rawpk %>%
  ggplot() +
  geom_point(aes(x = TAD, y = DV, color = ID, shape = IND)) +
  geom_line(aes(x = TAD, y = DV, group=interaction(ID, IND), color = ID, linetype = IND)) +
  labs(
    x ="Time after dose (h)", 
    y = "Concentration (ng/ml)", 
    title ="Midazolam plasma concentration", 
    color = ""
  ) +
  theme_classic() +
  theme(
    plot.title = element_text(hjust = 0.5), 
    legend.position = "bottom", 
    legend.title = element_blank(), 
    text=element_text(family="Times")
  )

# Individual plots of midazolam plasma concentration and time.
ggplot(data = rawpk, aes(x = TAD, y = DV, color =IND)) +
  geom_point() +
  facet_wrap(~ID) +
  scale_y_log10() +
  labs(
    x ="Time after dose (h)", 
    y = "Concentration (ng/ml)", 
    title ="Midazolam plasma concentration (log-transformed)"
  ) + 
  theme(axis.text.x = element_text(angle = 90), text=element_text(family="Times"))

```


## Section VI: Paired t-testing

```{r}
# Calculate mean AUC_last +/- standard deviation (SD) for the two groups (no inducer/inducer).
pk_analysis %>% 
  group_by(IND) %>% 
  summarise(mean(AUClast), SD = sd(AUClast))

# Q-Q plot to check for normality. 
ggqqplot(pk_analysis$AUClast)

# If normality is not satisfied, log-transform the AUC_last values.
pk_analysis$logAUClast = log(pk_analysis$AUClast)

# Compute the paired t-test for AUC_last.
t.test(pk_analysis$AUClast, pk_analysis$IND, paired = TRUE)
```


## Section VII: Sample size calculation

```{r}
# SD_pooled = sqrt((SD_1^2 + SD_2^2)/2) = sqrt((1.12^2 + 1.09^2)⁄2)) = 1.11
# Cohen's d = (mean_2-mean_1)/SD_pooled = (4.47 - 6.33)/1.11 = -1.68
pwr.t.test(
  n= NULL, d=-1.68, sig.level = 0.05, power =0.80, type = c("paired"), alt = c("two.sided") 
  )        

```

